---
title: "Vote By Mail Analysis"
author: "David F. Severski (@dseverski)"
output:
  html_notebook: 
    theme: cerulean
---

Inspired by dinner-table speculation on what point in the 2016 election 
cycle the actual vote outcome was determined (early voting, absentee votes, 
states that close ahead of others, etc.), I was interested in looking at 
rates of absentee voting in the US. My household have been permanent vote-by-mail 
voters for years and it's hard for me to envision actually going to a polling 
location. __[Whacky US Election Laws...*sigh*]--

A quick web searched turned up [this](http://www.nytimes.com/interactive/2012/10/07/us/voting-by-mail.html) 
graphic from the NYTimes, which referenced the US Election Assistance 
Commission (EAC) as a data source.

Data available at:  https://www.eac.gov/research/election_administration_and_voting_survey.aspx
Code book: https://www.eac.gov/assets/1/Page/2014%20EAVS%20Data%20File%20Codebook-2015-06-24-V03-Final.pdf


```{r setup}
library(haven)      # for SPSS data file imports
library(tidyverse)  # for data processing assistance
library(viridis)    # for our color scheme
```

The EAC makes data available in several formats. The SPSS dataset looks to be 
the most ammenable. Below we download the ZIP file, expand it, and read in 
Section C data (absentee info).

```{r fetch_data}
url <- "https://www.eac.gov/assets/1/Page/2014_EAVS_SPSS_Files.zip"
if (!file.exists(file.path("data", "2014_EAVS_SPSS_Files.zip"))) {
  download.file(url, "data/2014_EAVS_SPSS_Files.zip")
}
unzip(file.path("data", "2014_EAVS_SPSS_Files.zip"), exdir = "data", overwrite = TRUE)
dat <- haven::read_sav("data/EAVS_Section_C.sav")
dat
```

```{r explore_data}
#class(dat)
dat %>% group_by(State) %>% summarize(total_absentee = sum(QC4_Total, na.rm = TRUE)) %>% 
  arrange(desc(total_absentee))
table(dat$QC2)
dat %>% group_by(State, QC2) %>% tally
```

I'm curious on region trends. Combine the EAC data with the built in R datasets 
on state regions.

```{r enhance_with_regions}
state_regions <- data_frame(state.name = state.name, state.abb = state.abb, region = as.character(state.region))
data_with_regions <- left_join(dat, state_regions, by = c("State" = "state.abb"))
data_with_regions %>% group_by(region) %>% tally(QC4_Total)
```

```{r get_total_voters}
total_votes <- haven::read_sav("data/EAVS_Section_A.sav")
total_votes_by_state <- group_by(total_votes, FIPSCode, PreferredOrder) %>% 
  summarize(registered_voters = sum(QA1a, na.rm = TRUE))
total_votes_by_state
```


```{r percentage_absentee}
voters_absentee <- left_join(data_with_regions, total_votes_by_state)
#dat2 <- group_by(voters_absentee, region, State, FIPSCode) %>% 
dat2 <- group_by(voters_absentee, State, state.name) %>% 
  summarize(percent_absentee = sum(QC4_Total, na.rm = TRUE) / sum(registered_voters) * 100) %>% 
  arrange(desc(percent_absentee)) %>% 
  filter(percent_absentee != Inf)
dat2
```

The following graph could be greatly improved -- spatial analysis isn't my 
forte. Immediate improvements include moving this down to the FIPS code 
and improving the breaks (it's not clear that OR has a higher pct than either WA 
or CO).

```{r graph_it}
state_map <- map_data("state")
gg <- ggplot()
gg <- gg + geom_map(data = state_map, map = state_map, 
                    aes(long, lat, map_id = region))
gg <- gg + geom_map(data = dat2, map = state_map,
                    aes(fill = percent_absentee, 
                        map_id = tolower(state.name)))
#gg <- gg + scale_fill_continuous(low='thistle2', high='darkred',  guide='colorbar')
gg <- gg + labs(title = "Percentage absentee voters vs. all registered voters by state",
                caption = "Data: US EAC - 2014 General Election",
                x = NULL, y = NULL,
                fill = "Pct Absentee of Total")
gg <- gg + coord_map("polyconic") 
gg <- gg + theme_minimal() + theme(legend.position = "bottom")
#gg <- gg + guide_legend(title = "Percent Absentee of Total")
gg <- gg + scale_fill_viridis(alpha = 0.7, guide = guide_legend(position = "bottom", direction = "horizontal"))
gg
```

There's a lot of great looking data in the EAC data. More exploration is 
warranted!

